---
title: "Retail Strategy and Analytics"
mainfont: Roboto
monofont: Consolas
output:
  pdf_document:
    df_print: default
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```
```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
# this hook is used only when the linewidth option is not NULL
if (!is.null(n <- options$linewidth))
{
x = knitr:::split_lines(x)
# any lines wider than n should be wrapped
if (any(nchar(x) > n))
x = strwrap(x, width = n)
x = paste(x, collapse = "\n")
}
hook_output(x, options)
})
```

## Load required libraries and datasets
```{r 0 Load Libraries}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library(data.table)
library(ggplot2)
library(ggmosaic)
library(readr)

#### Point the filePath to where you have downloaded the datasets to and
#### assign the data files to data.tables

filePath <- "C:/New D/DA Projects/Project 4/"
transactionData <- fread(paste0(filePath, "QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath, "QVI_purchase_behaviour.csv"))

```
## Exploratory data analysis
The first step in any analysis is to first understand the data. Let's take a look
at each of the datasets provided.
### Examining transaction data
```{r Examining transaction data}
#### Examine transaction data
str(transactionData)

#### Convert DATE column to a date format
# Assuming the "DATE" column represents days since an origin (e.g., Excel's date format)
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30") 
```
```{r summary of PROD_NAME}
#### Examine PROD_NAME
transactionData[,.N,PROD_NAME]
```
```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
productWords <- data.table(unlist(strsplit(unique(transactionData[,PROD_NAME])," ")))
setnames(productWords,'words')
```
```{r}
### Removing digits
productWords <- productWords[grepl("\\d", words) == FALSE, ]
### Removing Special Characters
productWords<-productWords[grepl("[:alpha:]",words), ]
#### Let's look at the most common words by counting the number of times a word appears and
#### sorting them by this frequency in order of highest to lowest frequency
productWords[,.N,words][order(-N)]
```
```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```
```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
summary(transactionData)
```
```{r }
#### Filter the dataset to find the outlier
transactionData[PROD_QTY==200, ]
```
```{r}
#### Let's see if the customer has had other transactions
transactionData[LYLTY_CARD_NBR==226000, ]
```
It looks like this customer has only had the two transactions over the year and is
not an ordinary retail customer. The customer might be buying chips for commercial
purposes instead. We'll remove this loyalty card number from further analysis.
```{r}
#### Filter out the customer based on the loyalty card number
transactionData<-transactionData[LYLTY_CARD_NBR!=226000, ]
#### Re-examine the transaction data
summary(transactionData)
```
let's look at the number of transaction lines over time to see
if there are any obvious data issues such as missing data.
```{r}
#### Count the number of transactions by date
transactionData[ ,.N,by=DATE]
```
There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's
create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a
chart of number of transactions over time to find the missing date.
```{r fig.align = "center"}
#### Create a sequence of dates and join this the count of transactions by date
allDates <- data.table(seq(as.Date("2018-07-01"), as.Date("2019-06-30"), by =
"day"))

setnames(allDates,"DATE")
transactions_by_day <- merge(allDates,transactionData[ ,.N,by=DATE],all.x=TRUE)

#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#### Plot transactions over time
ggplot(transactions_by_day,aes(x=DATE,y=N)) + geom_line() + labs(x="Day",y="No. of transactions",title="Transactions over time") + scale_x_date(breaks="1 month") +
 theme(axis.text.x = element_text(angle=90,vjust=0.5))


```
We can see that there is an increase in purchases in December and a break in late
December. Let's zoom in on this.
```{r fig.align = "center"}
#### Filter to December and look at individual days
ggplot(transactions_by_day[month(DATE)==12, ], aes(x=DATE,y=N)) + geom_line()+
  labs(x="Day",y="No. of transactions",title="Transactions over time") +
  scale_x_date(breaks="1 day")+ theme(axis.text.x = element_text(angle=90,vjust=0.5))
```
We can see that the increase in sales occurs in the lead-up to Christmas and that
there are zero sales on Christmas day itself. This is due to shops being closed on holiday
There are no more outliers
Lets creature extra features such as pack size and brand name  from PROD_NAME column
```{r Create pack size}
#### Pack size
#### We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]
####  checking if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
#### Checking the updated data
transactionData
```
```{r}
#### Creating histogram of PACK_SIZE
hist(transactionData[,PACK_SIZE])

```
```{r Create brand name}
#### Brands

transactionData[,BRAND := toupper(substr(PROD_NAME,1,regexpr(pattern=' ',PROD_NAME)-1))]
#### Checking brands
transactionData[,.N,by=BRAND][order(-N)]

```
Some of the brand names look like they are of the same brands - such as RED and
RRD, which are both Red Rock Deli chips. Let's combine these together.
```{r Clean brand names}
#### Combine brand names

transactionData[BRAND == "RED", BRAND := "RRD"]
transactionData[BRAND == "SNBTS", BRAND := "SUNBITES"]
transactionData[BRAND == "INFZNS", BRAND := "INFUZIONS"]
transactionData[BRAND == "WW", BRAND := "WOOLWORTHS"]
transactionData[BRAND == "SMITH", BRAND := "SMITHS"]
transactionData[BRAND == "NCC", BRAND := "NATURAL"]
transactionData[BRAND == "DORITO", BRAND := "DORITOS"]
transactionData[BRAND == "GRAIN", BRAND := "GRNWVES"]

#### Check again
transactionData[, .N, by = BRAND][order(BRAND)]
```
### Examining customer data
```{r 1 Exploratory data analysis}
#### Examining customer data
str(customerData)

summary(customerData)
```
Let’s have a closer look at the LIFESTAGE and PREMIUM_CUSTOMER columns.
```{r}
#### Examining the values of lifestage and premium_customer
customerData[,.N,by=LIFESTAGE][order(-N)]

customerData[,.N,by=PREMIUM_CUSTOMER][order(-N)]

```
Joining the transaction and customer datasets together
```{r }
#### Merge transaction data to customer data
data<-merge(transactionData,customerData,all.x=TRUE)
```
As the number of rows in `data` is the same as that of `transactionData`, we can be
sure that no duplicates were created. This is because we created `data` by setting
`all.x = TRUE` (in other words, a left join) which means take all the rows in
`transactionData` and find rows with matching values in shared columns and then
joining the details in these rows to the `x` or the first mentioned table
```{r Check for missing customer details}
data[is.null(LIFESTAGE),.N]

data[is.null(PREMIUM_CUSTOMER),.N]
```
#Data Exploration completed
```{r Code to save dataset as a csv}
fwrite(data, paste0(filePath,"QVI_data.csv"))
```
## Data analysis on customer segments 
Defining Metrics

- Who spends the most on chips (total sales), describing customers by lifestage and
how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment

Calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and 
plotting the split by these segments to describe which customer segment contribute 
most to chip sales.
```{r fig.width = 10, fig.align = "center"}
#### Total Sales by LIFESTAGE AND PREMIUM_CUSTOMER
sales<- data[,.(SALES=sum(TOT_SALES)),.(LIFESTAGE,PREMIUM_CUSTOMER)]
#### Create plot
pl <-ggplot(data=sales)+geom_mosaic(aes(weight=SALES,x=product(PREMIUM_CUSTOMER,LIFESTAGE),        fill=PREMIUM_CUSTOMER))+ labs(x="Lifestage",y="Premium customer flag",title="Proportion of sales") + theme(axis.text.x = element_text(angle=90,vjust=0.5))
#### Plot and label with proportion of sales
pl + geom_text(
  data = ggplot_build(pl)$data[[1]],
  aes(
    x = (xmin + xmax) / 2,
    y = (ymin + ymax) / 2,
    label = as.character(paste(round(.wt / sum(.wt), 3) * 100, '%'))
  )
)

```
Sales are coming mainly from Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees
Let's see if the higher sales are due to there being more customers who buy chips.
```{r fig.width = 10, fig.align = "center"}
#### Number of customers by LIFESTAGE and PREMIUM_CUSTOMER
customers<- data[,.(CUSTOMERS=uniqueN(LYLTY_CARD_NBR)),
                 .(LIFESTAGE,PREMIUM_CUSTOMER)][order(-CUSTOMERS)]

#### Create plot
p <- ggplot(data=customers)+
     geom_mosaic(aes(weight=CUSTOMERS,x=product(PREMIUM_CUSTOMER,LIFESTAGE),
                                             fill=PREMIUM_CUSTOMER)) +
     labs(x="Lifestage",y="Premium customer flag",title="Proportion of customers") +
      theme(axis.text.x =element_text(angle=90,vjust=0.5))

#### Plot and label with proportion of customers
p + geom_text(
  data = ggplot_build(p)$data[[1]],
  aes(
    x = (xmin + xmax) / 2,
    y = (ymin + ymax) / 2,
    label = as.character(paste(round(.wt / sum(.wt), 3) * 100, '%'))
  )
)
```
There are more Mainstream - young singles/couples and Mainstream - retirees who buy
chips. This contributes to there being more sales to these customer segments but 
this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer. 
Let's have a look at this next.

```{r fig.width = 10, fig.align = "center"}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
avg_units<-data[,.(AVG=sum(PROD_QTY)/uniqueN(LYLTY_CARD_NBR)),
                .(LIFESTAGE,PREMIUM_CUSTOMER)][order(-AVG)]

#### Create plot
ggplot(data=avg_units,aes(weight=AVG,x=LIFESTAGE,fill=PREMIUM_CUSTOMER)) +
  geom_bar(position= position_dodge()) +
  labs(x="Lifestage",y="Avg units per transaction",  title="Units per customer")+
  theme(axis.text.x = element_text(angle=90,vjust=0.5))
  

```
Older families and young families in general buy more chips per customer

Let's find out the average price per unit chips bought for each customer 
segment as this is also a driver of total sales.
```{r fig.width = 10, fig.align = "center"}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER
avg_price<-data[,.(AVG=sum(TOT_SALES)/sum(PROD_QTY)),
                .(LIFESTAGE,PREMIUM_CUSTOMER)] [order(-AVG)]

#### Create plot
ggplot(data=avg_price,aes(weight=AVG,x=LIFESTAGE,fill=PREMIUM_CUSTOMER)) +
      geom_bar(position=position_dodge()) +
       labs(x="Lifestage",y="Avg price per unit",title="Price per unit") +
       theme(axis.text.x = element_text(angle=90,vjust=0.5))


```
Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to
buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than their own
consumption. This is also supported by there being fewer premium midage and young singles and couples
buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn’t large, we can check if this difference is statistically different.
```{r}
#### Perform an independent t-test between mainstream vs premium and budget midage and
#### young singles and couples
PricePerUnit<-data[,price:=TOT_SALES/PROD_QTY]

t.test(data[LIFESTAGE%in%c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES")
       & PREMIUM_CUSTOMER == "Mainstream",price]
       ,data[LIFESTAGE%in%c("YOUNG SINGLES/COUPLES","MIDAGE SINGLES/COUPLES")
       & PREMIUM_CUSTOMER != "Mainstream",price]
       , alternative="greater")
      

```
The t-test results in a p-value < 2.2e-16, i.e. the unit price for mainstream, young and mid-age singles and couples are significantly higher than that of budget or premium, young and midage singles and couples

## Deep dive into specific customer segments for insights
We have found  few interesting insights that we can dive deeper into.
We might want to target customer segments that contribute the most to sales to retain them or further increase sales. 

Let’s look at Mainstream - young singles/couples. For instance, let’s find out if they tend to buy a particular brand of chips.
```{r fig.align = "center"}
#### Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE=="YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER=="Mainstream",]
other <- data[!(LIFESTAGE=="YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER=="Mainstream"),]

#### Brand affinity compared to the rest of the population
quantity_segment1<-segment1[,sum(PROD_QTY)]
quantity_other<-other[,sum(PROD_QTY)]

quantity_segment1_by_brand<-  segment1[,.(targetSegment=sum(PROD_QTY)/quantity_segment1),by=BRAND]
quantity_other_by_brand <- other[, .(other=sum(PROD_QTY)/quantity_other),by=BRAND]

brand_proportions<-merge(quantity_segment1_by_brand,
                         quantity_other_by_brand)[,affinityToBrand:=targetSegment/other]
         

brand_proportions[order(-affinityToBrand)]
```
We can see that :
• Mainstream young singles/couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population
• Mainstream young singles/couples are 56% less likely to purchase Burger Rings compared to the rest
of the population.

Checking if our target segment tends to buy larger packs of chips.

```{r fig.align = "center"}
#### Preferred pack size compared to the rest of the population
quantity_segment1_by_pack<-segment1[, .(targetSegment=sum(PROD_QTY)/quantity_segment1),
                                     by=PACK_SIZE]
quantity_other_by_pack<-other[, .(other=sum(PROD_QTY)/quantity_other),by=PACK_SIZE]

pack_proportions<-merge(quantity_segment1_by_pack,
                        quantity_other_by_pack)[,affinityToPack:= targetSegment/other]
pack_proportions[order(-affinityToPack)]

```
It looks like Mainstream young singles/couples are 27% more likely to purchase a 270g pack of chips compared to the rest of the population but let’s dive into what brands sell this pack size

Check for brands selling the particular PACK_SIZE

```{r}
data[PACK_SIZE==270,unique(PROD_NAME)]

```
Twisties are the only brand offering 270g packs and so this may instead be reflecting a higher likelihood of purchasing Twisties.

## Conclusion
1) Sales have mainly been due to Budget - older families, Mainstream - young singles/couples, and Mainstream- retirees shoppers.
2) We found that the high spend in chips for mainstream young singles/couples and retirees is due to there being more of them than other buyers.
3) Mainstream, midage and young singles and couples are also more likely to pay more per packet of chips. This is indicative of impulse buying behaviour.
4) We’ve also found that Mainstream young singles and couples are 23% more likely to purchase Tyrrells chips compared to the rest of the population. The Category Manager may want to increase the category’s performance by off-locating some Tyrrells and smaller packs of chips in discretionary space near segments where young singles and couples frequent more often to increase visibilty and impulse behaviour.
